---
title: "Project 3"
author: "Jasmine Gallaway and Keren Vivas"
date: "2023-10-30"
output: github_document
params:
 Education: "None/Elementary"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. Introduction section

The data analyzed in this report is derived from the Diabetes Health Indicators Dataset, which was gathered through the Behavioral Risk Factor Surveillance System (BRFSS) in the year 2015. The BRFSS is an annual health-related telephone survey conducted by the Centers for Disease Control and Prevention (CDC). This dataset encompasses a substantial collection of over 200,000 survey responses, featuring 21 distinct variables or factors, along with a pivotal response variable termed "Diabetes_binary." This response variable comprises three discrete levels:

0 = No diabetes or diabetes only during pregnancy.
1 = Prediabetes.
2 = Established diabetes.

The dataset encompasses a diverse array of variables that cover a wide spectrum of information, including the current diabetes status of the survey respondents, various demographic details, physical and mental health metrics, dietary habits, and lifestyle factors. The specific variables selected for analysis in this report are as follows:

- Demographics: This category includes two categorical variables: "Age" and "Sex" of the respondents.
- Physical Conditions: These variables encompass "HighBP," "HighChol," and "HeartDiseaseorAttack." These are categorical variables, taking values of 0 or 1, indicating the presence or absence of high blood pressure, high cholesterol, and a history of heart disease or heart attacks. Additionally, the "Body Mass Index (BMI)" is included as a continuous variable, with values ranging from 12 to 98, providing insights into the respondents' body composition.
- Mental Conditions: The variable "MentHlth" is a continuous variable that quantifies the number of days in a month during which the respondent's mental health was not good. It ranges from 1 to 30.
- Diet Conditions: Two categorical variables, "Fruits" and "Veggies," denote whether respondents consume fruits and vegetables one or more times per day. These variables take values of 1 (indicating consumption) or 0 (indicating non-consumption).
- Habits: This category includes two categorical variables: "Smoker" and "PhysActivity." "Smoker" indicates whether respondents have smoked at least 100 cigarettes in their lifetime (with values 1 for yes and 0 for no), while "PhysActivity" assesses whether respondents have engaged in physical activity within the last month (with values 1 for yes and 0 for no).

The Exploratory Data Analysis (EDA) of the selected variables is designed to offer a holistic understanding of the health and lifestyle characteristics of the survey respondents. This preliminary phase of analysis helps us uncover patterns within the data.

Subsequently, when we move to modeling the data, our objective is to delve deeper and gain insights into the intricate ways these factors impact an individual's diabetes status. Our analysis will not only reveal the influence of these factors but also quantify the individual contributions of each factor to the risk of diabetes.

In the end, the entire analysis will be segmented based on the level of education. This segmentation enables us to investigate how this specific demographic factor influences the overall findings. By doing so, we can discern whether educational levels play a significant role in shaping the relationship between health, lifestyle, and diabetes risk.

# 2. Required packages section
```{r required_packages}
library(readr)  
library(dplyr)  
library(ggplot2)  
library(caret) 
library(rmarkdown)
library(shiny)
library(purrr)  
library(randomForest)  
library(gbm)  
library(MLmetrics)  #For logLoss()
```
# 3. Data section
```{r cleaning_data}
#readin data  
diabetes_data <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")  

#making a tibble
diabetes_data <- as_tibble(diabetes_data)  

#changing diabetes_binary to factor  
diabetes_data$Diabetes_binary <- diabetes_data$Diabetes_binary %>% 
  factor(levels = c(0,1,2),
         labels = c("No Diabetes", "Prediabetes", "Diabetes"))  

#combine education levels one and two  
diabetes_data$Education <- diabetes_data$Education %>% 
  factor(levels = c(1,2,3,4,5,6), 
            labels = c("None/Elementary", "None/Elementary", "Some High School", "High School Graduate", "Some College", "College Graduate"))  

#changing sex to factor  
diabetes_data$Sex <- diabetes_data$Sex %>% 
  factor(levels = c(0,1),
         labels = c("female", "male"))  

#changing age to factor  
diabetes_data$Age <- diabetes_data$Age %>% 
  factor(levels = c(1,2,3,4,5,6,7,8,9,10,11,12,13),
         labels = c("18-24", "25-30", "31-35", "36-40", "41-45", "46-50", "51-55", "56-60", "61-65", "66-70", "71-75", "76-80", "80+"))  

#Changing fruit to factor  
diabetes_data$Fruits <- diabetes_data$Fruits %>% 
  factor(levels = c(0, 1),
         labels = c("None", "Yes 1 or more"))  

#Changing physical activity to factor  
diabetes_data$PhysActivity <- diabetes_data$PhysActivity %>% 
  factor(levels = c(0, 1),
         labels = c("None", "Some in past 30 days"))  

#Changing smoker to factor  
diabetes_data$Smoker <- diabetes_data$Smoker %>% 
  factor(levels = c(0, 1),
         labels = c("No", "Yes, has smoked at least 100 cigarettes"))  

#Changing stroke to factor  
diabetes_data$Stroke <- diabetes_data$Stroke %>%
  factor(levels = c(0, 1),
         labels = c("No", "Yes"))  

#Changing heart disease or attack to factor  
diabetes_data$HeartDiseaseorAttack <- diabetes_data$HeartDiseaseorAttack %>% 
  factor(levels = c(0, 1),
         labels = c("No", "Yes"))  

#Changing veggies to factor  
diabetes_data$Veggies <- diabetes_data$Veggies %>% 
  factor(levels = c(0, 1),
         labels = c("None", "Yes 1 or more"))  

#Changing Heavy alcohol consumption to factor  
diabetes_data$HvyAlcoholConsump <- diabetes_data$HvyAlcoholConsump %>% 
  factor(levels = c(0, 1),
         labels = c("No", "Yes"))  

#Changing any healthcare to factor  
diabetes_data$AnyHealthcare <- diabetes_data$AnyHealthcare %>% 
  factor(levels = c(0, 1),
         labels = c("No", "Yes"))  

#Changing nodocbcCost to factor  
diabetes_data$NoDocbcCost <- diabetes_data$NoDocbcCost %>% 
  factor(levels = c(0, 1),
         labels = c("No", "Yes"))  

#Changing GenHlth to factor  
diabetes_data$GenHlth <- diabetes_data$GenHlth %>% 
  factor(levels = c(1, 2, 3, 4, 5),
         labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))  

#Changing DiffWalk to factor  
diabetes_data$DiffWalk <- diabetes_data$DiffWalk %>% 
  factor(levels = c(0, 1),
         labels = c("No", "Yes")) 

#changing HighBP to factor
diabetes_data$HighBP <- diabetes_data$HighBP %>%
  factor(levels = c(0, 1),
         labels = c("no high BP", "high BP"))

#changing HighChol to factor
diabetes_data$HighChol <- diabetes_data$HighChol %>%
  factor(levels = c(0,1),
         labels = c("no high cholesterol", "high cholesterol"))

#changing CholCheck to factor
diabetes_data$CholCheck <- diabetes_data$CholCheck %>%
  factor(levels = c(0,1),
         labels = c("no check in 5 years", "check in 5 years"))

#changing Income to factor
diabetes_data$Income <- diabetes_data$Income %>%
  factor(levels = c(1,2,3,4,5,6,7,8),
         labels = c("Less than $10k", "$10k to less than $15k", "$15kto less than $20k", "$20k to less than $25k", "$25k to less than $35k", "$35k to less than $50k", "$50k to less than $75k", "$75k or more"))
``` 
```{r filter data for the specified education level}
# Subset data into a single Education level
subset_data <- diabetes_data %>%
 filter(Education == params$Education)
```

# 4. Summarizations section
```{r EDA}
#gathering summary statistics  
sum_data <- diabetes_data %>% 
  summarize(mean = mean(BMI),
            sd = sd(BMI),
            min = min(BMI),
            max = max(BMI),
            IQR = IQR(BMI))  
#printing out  
print(sum_data)  

#boxplot to visualize the summary values  
sum_plot1 <- ggplot(diabetes_data, aes(x = Age, y = BMI)) + 
  geom_jitter()  

sum_plot1  
```
```{r Contingency tables}

 #1-way contingency table
 one_way_table <- table(diabetes_data$Fruits)
 one_way_table
 
 #2-way contingency table
 two_way_table <- table(diabetes_data$Smoker, diabetes_data$HighChol)
 two_way_table
 
 #3-way contingency table
 three_way_table <- table(diabetes_data$HighBP, diabetes_data$HighChol, diabetes_data$HeartDiseaseorAttack)
 three_way_table
```

```{r Plots}
#Violin plot
ggplot(diabetes_data, aes(x = Sex, y = MentHlth)) +
  geom_violin(fill = "lightblue", color = "blue", alpha = 0.7) +
  labs(title = "Mental health condition across gender", x = "Gender", y = "Mental health condition") +
  theme_minimal()

#Bar plot
ggplot(data = diabetes_data, aes(x = Age, y = PhysActivity , fill = Age)) +
  geom_bar(stat = "identity") +
  labs(title = "Stacked Bar Plot of Physical activity days by Age Group", x = "Age Group", y = "Physical activity (days)") +
  theme_minimal()

#Scatter plot
ggplot(diabetes_data, aes(x = BMI, y = MentHlth)) +
  geom_point(aes(color = Sex), size = 3) +
  labs(title = "Scatterplot of BMI vs. Mental health", x = "BMI", y = "MentHlth") +
  theme_minimal()
```

# 5. Modeling
```{r Split data}
set.seed(3033)
intrain <- createDataPartition(y = diabetes_data$Diabetes_binary, p= 0.7, list = FALSE)
training <- diabetes_data[intrain,]
testing <- diabetes_data[-intrain,]
```
## _What log loss is and why we may prefer it to things like accuracy?_

Logarithmic loss is metric used to evaluate how well a model fits. Also known as cross-entropy loss, log loss is used on binary classification models where the responses are either 0 or 1. Log loss calculates the difference between the predicted probabilities and the actual values. The lower the log loss value, the better the fit of the model is to the data. Log loss is preferred with this data set because of our focus on the binary response of whether or not someone has diabetes. Log loss focuses on models that predict the class better and penalizes models that are further from the class. 

## _What a logistic regression is and why we apply it to this kind of data?_

Logistic regression is a form of predictive analysis in which the regression models the relationship(s) of a binary response variable to one or more other variables. Logistic regression is similar to the use of linear regression except that logistic regression is used for classification problems, has a range of 0 to 1 like log loss, and does not require a linear relationship between the variables.

## Fit  3 candidate logistic regression models and choose the best
```{r Three candidate logistic models}
# Model 1: Logistic Regression with default parameters
model1 <- glm(Diabetes_binary ~ ., data = training, family = binomial)
y_pred1 <- predict(model1, newdata = testing, type = "response")
log_loss_model1 <- logLoss(obs = testing$Diabetes_binary, pred = y_pred1)

# Model 2: Logistic Regression with different parameters (e.g., regularization)
model2 <- glm(Diabetes_binary ~ ., data = training, family = binomial, control = glm.control(alpha = 0.1))
y_pred2 <- predict(model2, newdata = testing, type = "response")
log_loss_model2 <- logLoss(obs = testing$Diabetes_binary, pred = y_pred2)

# Model 3: Logistic Regression with more parameter variations
model3 <- glm(Diabetes_binary ~ ., data = training, family = binomial, control = glm.control(alpha = 1.0, weight = TRUE))
y_pred3 <- predict(model3, newdata = testing, type = "response")
log_loss_model3 <- logLoss(obs = testing$Diabetes_binary, pred = y_pred3)

# Choose the model with the lowest log loss
log_loss_values <- c(log_loss_model1, log_loss_model2, log_loss_model3)
best_model_index <- which.min(log_loss_values)

best_model <- switch(
  best_model_index,
  model1 = model1,
  model2 = model2,
  model3 = model3
)
best_log_loss <- log_loss_values[best_model_index]
cat("The best model is Model", best_model_index, "with a log loss of", best_log_loss, "\n")
```

## _What a LASSO logistic regression model is and why we might try to use this over basic logistic regresion?_

LASSO (Least Absolute Shrinkage and Selection Operator) is a type of regularized logistic regression. It adds a penalty term to the logistic regression's likelihood function to encourage feature selection by driving some coefficients to exactly zero. This can improve model interpretability and reduce overfitting and that is why it is used over the basic logistic regresion.

LASSO logistic regression is a valuable tool when you have high-dimensional datasets with potentially redundant or irrelevant features. It can improve model performance, reduce overfitting, and enhance model interpretability by automatically selecting the most relevant predictors while eliminating the least important ones. However, the choice between basic logistic regression and LASSO logistic regression depends on the specific characteristics of your data and the goals of your analysis

## Fit a LASSO logistic regression model and choose the best model

model_lasso <- train(Diabetes_binary ~ ., data = training_data, method = "glmnet",
                     trControl = trainControl(method = "cv", number = 5),
                     tuneGrid = expand.grid(alpha = 1, lambda = seq(0.001, 1, by = 0.001)))


## _what a classification tree model is and why we might try to use it?_

A classification tree is a decision tree that is used for classification tasks. It recursively splits the data into subsets based on the predictors and their values to make predictions. Different values of the complexity parameter control the tree's depth and complexity. Cross-validation is typically used to select the best complexity parameter.

## Fit a classification tree with varying values for the complexity parameter and choose the best model (best complexity parameter)

library(rpart)

# Fit a classification tree
tree_model <- rpart(Diabetes_binary ~ ., data = training_data, method = "class")

# Define the control parameters for the rpart function
ctrl <- rpart.control(cp = 0.01) # Start with a small value for cp

# Fit the tree with cross-validation and find the best cp value
cv_tree_model <- rpart(Diabetes_binary ~ ., data = training_data, method = "class", control = ctrl)

# Print the complexity parameter table
printcp(cv_tree_model)

#Prune the tree
best_tree_model <- prune(cv_tree_model, cp = best_cp_value)


## _what a random forest is and why we might use it instead of a basic classification tree?_

A random forest is an ensemble tree method of modeling. Random forest is similar to the bagging method of modeling except that a random forest does not use all the variables/predictors. A random forest uses a subset of predictors which reduces variance better than bagging or a basic classification tree does. 

## Fit a random forest model and choose the best model

Add answer

## Two models thast was not done in class

Model 1: Add answer
Model 2: Add answer

## What the Model 1 is?

Add answer

## Fit a Model 1 and choose the best model

Add answer

## What the Model 2 is?

Add answer

## Fit a Model 2 and choose the best model

Add answer


## 6. Final Model Selection (Compare the six best above and declare an overall winner)

Add answer


## NO SURE IN WHAT POINT WE SHOULD ADD THIS PIECE OF CODE :/

## X. Automation
```{r params set up}
#Do it for 1 first and then all level of education
rmarkdown::render("Project 3.Rmd", out_put_file = "None_Elementary.html", params = list(Education = "None/Elementary"))

#For all level of education
#get unique education levels
#Education_levels <- unique(diabetes_data$Education)

#create filenames
#output_file <- paste0(Education_levels, "html")

#create a list for each level of education with just the level name parameter
#params = lapply(Education_levels, FUN = function(x){list(Education = x)})

#Put into a data frame
#reports <- tibble(output_file, params)

# knit
#apply(reports, MARGIN = 1, FUN = function(x){ 
# render(input = "Project 3.Rmd", output_file = x[[1]], params = x[[2]])
#})
```

End report! :)
